---
title: "Cell type annotation of rice dataset"
output:
  output: github_document
---

```{r, warning=FALSE, message=FALSE}
# Load resources. Libraries, paths, functions, themes.
source(file.path("..", "..", "scripts", "config.R"))
```

```{r}
#Read merged dataset
merge_os_mock <- readRDS(OSA_DATA_PATH_MOCK)
merge_os_nema <- readRDS(OSA_DATA_PATH)
```

## Annotating rice mock dataset by cell type

We use first the mock samples from three biological replicates (R1, R2, R3) to manually annotate the clusters based on the expression of experimentally validated by RNA *in situ* hybridization cell type specific markers in rice. 


```{r}
DimPlot(merge_os_mock, group.by = "orig.ident")
```
First we perform an unsupervised clustering to select the best resolution to annotate the cell types.

```{r, message=FALSE}
merge_os_mock <- FindNeighbors(merge_os_mock, dims = 1:30, 
                               verbose = FALSE,reduction="pca")
merge_os_mock <- FindClusters(merge_os_mock, graph.name="SCT_snn",
                              resolution=seq(0.1,1.5,0.1))
```

```{r}
DimPlot(merge_os_mock, group.by = "SCT_snn_res.1.2", label = TRUE)

```

```{r}
#Dotplot with markers

#Load file with known validated markers
markers <- read.table(file.path("..","..","data","resources","04_Cell_type_rice",
                                "validated_markers.txt"),header=TRUE, sep="\t")
markers_vector <- c(markers$Gene)
markers_name <- c(markers$Cell_type)

markers_seurat <- setNames(markers_vector, markers_name)

p <-DotPlot(object = merge_os_mock, features = markers_seurat,
            group.by = "SCT_snn_res.1.2")

#Sort levels of cell types

markers_dotplot<-p$data %>% 
  mutate(feature.groups = factor(feature.groups,levels = c("Vascular initials",
                                                           "Dividing cells", 
                                                           "Sclerenchyma",
                                                           "Exodermis",
                                                           "Exo_Scl",
                                                           "Epidermis",
                                                           "Atrichoblast",
                                                           "Trichoblast",
                                                           "Cortex",                                                       
                                                           "Phloem",
                                                           "Endodermis",
                                                           "Pericycle_Endodermis",
                                                           "Pericycle",
                                                           "Xylem"))) %>% 
  arrange(factor(feature.groups, levels = c("Vascular initials",
                                            "Dividing cells", 
                                            "Sclerenchyma",
                                            "Exodermis",
                                            "Exo_Scl",
                                            "Epidermis",
                                            "Atrichoblast",
                                            "Trichoblast",
                                            "Cortex",                                                       
                                            "Phloem",
                                            "Endodermis",
                                            "Pericycle_Endodermis",
                                            "Pericycle",
                                            "Xylem"))) %>% 
  group_by(feature.groups) %>% 
  arrange(desc(avg.exp.scaled),group_by=TRUE)

#Sort clusters by avg expression data inthe dotplot (manual solution)
markers_dotplot <- markers_dotplot %>% 
  mutate(id=factor(id, levels = rev(c(14,20,8,12,15,25,4,6,18,21,1,9,2,19,13,
                                      24,5,11,16,10,7,0,22,17,3,23))))


dotplot_os_mock <- ggplot(data        = markers_dotplot,
       aes(y       = id,
           x       = features.plot,
           col     = avg.exp.scaled)) +
  geom_point(aes(size     = pct.exp))+
  scale_size_area(max_size = 2)+
  scale_colour_gradientn(colours=c("black", "magenta"))+
  facet_grid(
    cols = vars(feature.groups), drop = TRUE,scales = "free")+
  theme_minimal()+ theme(legend.position = "none",
                         text = element_text(size=10,family="sans"),
                         axis.text=element_text(size=10, family="sans"),
                         axis.title = element_blank(),
                         axis.text.x=element_text(angle=90),
                         element_line(linewidth = 0.1),
                         strip.text.x = 
                           element_text( margin = margin( b = 1, t = 6),
                                         size=10),
                         panel.border = 
                           element_rect(colour = "black", fill=NA, 
                                        linewidth = 0.1),
                         strip.background = 
                           element_rect(colour = "black", linewidth = 0.1),
                         panel.spacing = unit(0.01, "lines"))

```

```{r, fig.width= 10}
dotplot_os_mock
```

Now, we are going to visualize the expression of the markers validated in this paper by RNA ISH or FISH

```{r}
#FeaturePlot validated markers x6
plot_Os12g0569200<- FeaturePlot(merge_os_mock, features = "Os12g0569200",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")#+ggtitle(NULL)
plot_Os05g0438700<- FeaturePlot(merge_os_mock, features = "Os05g0438700",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")#+ggtitle(NULL)
plot_Os10g0454200<- FeaturePlot(merge_os_mock, features = "Os10g0454200",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")##+ggtitle(NULL)

plot_Os08g0115800<- FeaturePlot(merge_os_mock, features = "Os08g0115800",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")#+ggtitle(NULL)
plot_Os06g0681600<- FeaturePlot(merge_os_mock, features = "Os06g0681600",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")#+ggtitle(NULL)
plot_Os11g0539200<- FeaturePlot(merge_os_mock, features = "Os11g0539200",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")#+ggtitle(NULL)
plot_Os05g0597100<- FeaturePlot(merge_os_mock, features = "Os05g0597100",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")#+ggtitle(NULL)
plot_Os03g0130300<- FeaturePlot(merge_os_mock, features = "Os03g0130300",
                                pt.size = 0.05,order = TRUE)+
  scale_colour_gradientn(colours=c("black", "magenta"))+theme_void()+
  theme(legend.position="none")#+ggtitle(NULL)


```

```{r}
grid.arrange(plot_Os05g0438700, plot_Os06g0681600, plot_Os08g0115800,
             plot_Os10g0454200, plot_Os11g0539200, plot_Os12g0569200,
             plot_Os05g0597100,plot_Os03g0130300,
             ncol=3)
```

With this information we are going to annotate the clusters manually.

```{r}
Idents(merge_os_mock) <- "SCT_snn_res.1.2"
  
merge_os_mock<-RenameIdents(merge_os_mock,
                            "0"="Cortex",
                        "1"= "Epidermis",
                        "2"="Epidermis",
                        "3"="Pericycle",
                        "4"="Sclerenchyma",
                        "5"="Trichoblast",
                        "6"="Sclerenchyma",
                        "7"="Cortex",
                        "8"="Dividing",
                        "9"="Epidermis",
                        "10"="Cortex",
                        "11"="Trichoblast",
                        "12"="Sclerenchyma",
                        "13"="Atrichoblast",
                        "14"="VascInitials",
                        "15"="Exodermis",
                        "16"="Trichoblast",
                        "17"="Endodermis",
                      "18" = "Exodermis",
                      "19" = "Epidermis",
                      "20" = "Dividing",
                      "21" = "Epidermis",
                      "22" = "Phloem",
                      "23" = "Xylem",
                      "24" = "Atrichoblast",
                      "25" = "Exodermis")
merge_os_mock <-
  AddMetaData(merge_os_mock,Idents(merge_os_mock),col.name = "Cell_type")
```

Based on the RNA ISH of Os10g0454200 that it is expressed in the young epidermis and the root cap, we suspect that a subcluster of the cluster 20 could correspond to the root cap. We increase the resolution to be able to annotate this subcluster. 

```{r}
Idents(merge_os_mock)<-"SCT_snn_res.1.2"
merge_os_mock <- FindSubCluster(merge_os_mock, cluster=20, resolution=0.2, graph.name = "SCT_snn")
```

```{r}
DimPlot(merge_os_mock, group.by = "sub.cluster", label = TRUE)
```
We rename the cluster 20_1 as root cap (RC)

```{r}
merge_os_mock@meta.data<- merge_os_mock@meta.data %>% 
  mutate(Cell_type = case_when(sub.cluster == "20_1" ~ "RC",
                               TRUE ~ Cell_type))
Idents(merge_os_mock)<- "Cell_type"
```

```{r}
DimPlot(merge_os_mock, group.by = "Cell_type")
```

```{r}
#Save object annotated
#saveRDS(merge_os_mock,OSA_DATA_PATH_MOCK)
```

## Transferring cell type annotation to the complete dataset

```{r, eval=FALSE}
Idents(merge_os_mock) <- "Cell_type"

os_anchors <- FindTransferAnchors(reference=merge_os_mock, query=merge_os_nema, dims=1:30, reference.reduction = "pca")
new_labels <- TransferData(anchorset = os_anchors, refdata = Idents(merge_os_mock),
                           dims = 1:30)
```


```{r, include=FALSE}
new_labels <- readRDS(file.path("..","..", "data","resources",
                      "04_Cell_type_rice", 
                      "annotation.prediction.rds"))
```


```{r}
merge_os_nema <- AddMetaData(object =merge_os_nema, 
                             metadata = new_labels$predicted.id,
                             col.name = "predicted.id")#Add predicted id
merge_os_nema <- AddMetaData(object =merge_os_nema, 
                             metadata = new_labels$prediction.score.max,
                             col.name = "prediction.score.max")#Add score prediction
```

```{r}
DimPlot(merge_os_nema, group.by = "predicted.id")
```


```{r}
#We are going to annotate dividing and VascInit as Initials
merge_os_nema@meta.data <- merge_os_nema@meta.data %>% 
  mutate(Cell_type= case_when(predicted.id %in% c("VascInitials","Dividing") ~ "Initials", 
                            TRUE ~ predicted.id)) %>% 
  mutate(Cell_type = fct_relevel(Cell_type, c("RC","Initials","Epidermis","Atrichoblast",
                                          "Trichoblast", "Exodermis", "Sclerenchyma","Cortex",
                                          "Endodermis","Pericycle", "Phloem", "Xylem")))
```

```{r}
DimPlot(merge_os_nema, group.by = "Cell_type")
```

```{r}
sessionInfo()
```

